Type theories as quotient inductive-recursive types

### 2025-01-18 tex/main.tex
List formal definitions

\usepackage{todonotes}

@@

\begin{abstract}
  Type theory can be defined in a type theory as a quotient inductive-inductive type, but terms are painstakingly littered with explicit coercions, i.e.\ \transp's, along equality constructors.
  This mess has been dubbed the \emph{transport hell} by the community in private but typically suppressed in presentation for clarity, ignoring the elephant in the room.
  In this paper, we aim for shortening the gap between formal and informal definitions of type theory in type theory using quotient inductive-inductive-recursive types and definitions by overlapping patterns altogether, reducing the use of \transp's.
  As a case study, we investigate (parallel) substitution calculus and type theory with Coquand universes and $\Pi$-types as quotient inductive-inductive-recursive types with meta-theoretic properties proved and compare our definitions with quotient inductive-inductive definitions.
\end{abstract}

\section{Introduction} \label{sec:intro}
Points to elaborate:
\begin{enumerate}
  \item QIITs are littered with transports, causing too much pains while formalising a type theory.
  \item Rewrite rules are popular to make structural rules definitional but requires efforts to justify meta-theoretic properties in general.
  \item Further, the confluence check adopted by \Agda is based on complete development and requires one-step parallel reductions to have a confluent term, making a definition more complicated than necessary.
  \item Moreover, the use of rewrite rules is typically \emph{not} presented and discussed in literature. We are ignoring the elephant in the room.
  \item Instead, definitions by overlapping patterns (in theory) requires strong normalisation and local confluence, making definitions simpler to design.
  \item The rules for type substitutions are \emph{structural} shared with other type theories based on parallel substitution (in particular cwf).

\end{enumerate}

\paragraph*{Contributions}
\begin{itemize}
  \item Exploration of the use of quotient inductive-inductive-recursive types and definitions by overlapping patterns.
    In particular, we give in a type theory:
    \begin{itemize}
      \item a definition of parallel substitution calculus;
      \item a definition of type theory with Coquand universes and $\Pi$-types,
    \end{itemize}
   whose type substitution rules are definitional using quotient inductive-inductive-recursive types and definitions by overlapping patterns.
  \item Comparison with other definitions using quotient inductive-inductive types.
\end{itemize}

@@

\section{Metatheory and formalisation}
\cite{Pujet2022,Pujet2024,Pujet2022a}
Points to include:
\begin{enumerate}
  \item identity type $x =^{A} y$ for $x, y : A$, dependent identity type, $t =^{P}_{p} u \defeq \transfib{P}{p}{t} =^{P y} u$ for $t : P(x)$ and $u : P(y)$, heterogeneous equality in \Agda $x \simeq y$ for $x : A$ and $y : B$
  \item 
\end{enumerate}

@@

Mention that the recursion part of induction-recursion are already not defined by elimination rule.

\subsection{Definitions by overlapping patterns}
\cite{Cockx2014,Altenkirch2016a}
\subsection{Quotient inductive-inductive-recursive types}
\subsection{Definitions by rewrite rules}
\cite{Cockx2020,Cockx2021}

@@

\section{Type theories as quotient inductive-inductive-recursive types} \label{sec:QIIRTs}
\subsection{Parallel substitution calculus} \label{subsec:SC}
\begin{alignat*}{3}
  \Ctx   & : \Set                   \\
  \Ty    & : \Ctx \to \Set          \\
  \Sub   & : \Ctx \to \Ctx \to \Set \\
  \Tm    & : (\Gamma : \Ctx) \to \Ty\,\Gamma \to \Set
\end{alignat*}
\begin{alignat*}{3}
  \emptyctx & : \Ctx \\
  \blank,\blank & : (\Gamma : \Ctx) \to \Ty\,\Gamma \to \Ctx
\end{alignat*}
\begin{alignat*}{3}
  [\blank]\blank &: \implicit{\Gamma, \Delta} \; \Sub\,\Gamma\,\Delta \to \Ty\,\Delta \to \Ty\,\Gamma \\
  U & : \implicit{\Gamma}\;\Ty\, \Gamma
\end{alignat*}
\begin{alignat*}{3}
  \emptysub & : \implicit{\Gamma}\;\Sub\,\Gamma\,\emptyctx \\
  \blank,\blank & : \implicit{\Gamma, \Delta, A}\;(\sigma : \Sub\,\Gamma\,\Delta) \to \Tm\,\Gamma\,([ \sigma ]_{\Ty} A) \to \Sub\,\Gamma,(\Delta, A) \\
  \idS & : \implicit{\Gamma}\;\Sub\,\Gamma\,\Gamma \\
  \blank;\blank & : \implicit{\Gamma, \Delta, \Theta}\;\Sub\,\Gamma\,\Delta \to \Sub\,\Delta\,\Theta \to \Sub\,\Gamma\,\Theta \\
  \pi_1 & : \implicit{\Gamma, \Delta, A}\;\Sub\,\Gamma\,(\Delta, A) \to \Sub\,\Gamma\,\Delta \\
  \pi_2 & : \implicit{\Gamma, \Delta, A}\;(\sigma : \Sub\,\Gamma\,(\Delta, A)) \to \Tm\,\Gamma\,([ \pi_1\,\sigma ]\, A) \\
  [\blank] \blank & : \implicit{\Gamma,\Delta, A}\;(\sigma : \Sub\,\Gamma\,\Delta) \to \Tm\,\Delta\,A \to \Tm\,\Gamma\,([\sigma]\, A)
\end{alignat*}

@@

\begin{alignat*}{5}
  \mathsf{idr}    & : \implicit{\Gamma, \Delta, \sigma} && {\sigma ; \idS_{\Delta}} && =^{\Sub\,\Gamma\,\Delta} && {\sigma} \\
  \mathsf{idl}    & : \implicit{\Gamma, \Delta, \sigma} && {\idS_{\Gamma} ; \sigma} && =^{\Sub\,\Gamma\,\Delta} && {\sigma} \\
  ;\text{-}\mathsf{assoc} & : \implicit{\Gamma, \Delta, \Theta, \Xi, \sigma, \tau, \gamma} && (\sigma ; \tau) ; \gamma && =^{\Sub\,\Gamma\,\Theta} &&  \sigma ; (\tau ; \gamma) \\
  \mathsf{concat} & : \implicit{\Gamma, \Delta, \Theta, \sigma, \tau, A, t} && \sigma ; (\tau , t)      && =^{\Sub\,\Gamma\,(\Theta, A)} &&  (\sigma ; \tau) , \alert{\transfib{\Tm\,\Gamma}{[;]_{\Ty}^{-1}}{\color{black}[ \sigma ] t}} \\
  \emptyctx\eta   & : \implicit{\Gamma, \sigma} && \sigma                   && =^{\Sub\,\Gamma\,\emptyctx} & \emptysub \\
  \pi\eta         & : \implicit{\Gamma, \Delta, \sigma} && \sigma                   && =^{\Sub\,\Gamma\,(\Delta, A)} &&  (\pi_1 \sigma, \pi_2 \sigma) \\
  \pi_1\beta      & : \implicit{\Gamma, \Delta, \Theta, \sigma, A, t} && \pi_1(\sigma , t)        && =^{\Sub\,\Gamma,\Delta} &&  \sigma \\
  \pi_2\beta      & : \implicit{\Gamma, \Delta, \Theta, \sigma, A, t} && \pi_2(\sigma , t)        && =^{\Tm\,\Gamma\,([\blank]\,A)}_{\alert{\pi_1\beta}} &&  t \\
  [\idS]_t         & : \implicit{\Gamma, A, t} && {[\,\idS\,]\,t}          && =^{\Tm\,\Gamma}_{\alert{[\idS]_\Ty}}  && t \\
  [;]_t            & : \implicit{\Gamma, \Delta, \Theta, \sigma, \tau, t} && {[\,\sigma ; \tau\,]\,t} && =^{\Tm\,\Gamma}_{\alert{[;]_{\Ty}}}   && {[ \sigma ]\,[ \tau ]\,t} \\
\end{alignat*}

@@

  To better illustrate the annoyance, the complete proof of \eqref{eq:pi2-comp-real-proof} requires to show first three equations and the last equation explicitly by Lemma~2.3.4 and 2.3.2 in \cite{UFP2013} respectively:
  \begin{alignat*}{3}
         & ([\pi_1(\sigma; \tau)]\,A                    &&, \pi_2\,(\sigma ; \tau)) \label{eq:pi2-proof-1} \\
    = {} & ([\pi_1(\sigma; (\pi_1\tau , \pi_2\tau))]\,A &&, \pi_2\,(\sigma; (\pi_1\,\tau, \pi_2\,\tau))) \\
    = {} & ([\pi_1(\sigma;\pi_1\,\tau, t)]\,A &&, \pi_2\,(\sigma;\pi_1\,\tau, t)) \\
    = {} & ([\pi_1(\sigma;\pi_1\,\tau, t)]\,A &&, \pi_2\,(\sigma;\pi_1\,\tau, t)) \nonumber \\
    = {} & (\sub{\sigma;\pi_1\tau}{A} &&, [\sigma] (\pi_2\,\tau)) \nonumber
  \end{alignat*}
  where $p \judgeq [;]_{\Ty}^{-1} : \sub{\sigma}{\sub{\pi_1\,\tau}{A}} \to \sub{\sigma; \pi_1 \tau}{A}$ and $t \judgeq\transfib{\Tm\,\Gamma}{p}{[\sigma]\,(\pi_2\,\tau)}$.
\end{example}

\begin{alignat*}{3}
[ \idS ]_{\Ty}\,A             & \jdeq A \\
[ \sigma ; \tau ]_{\Ty}\,A    & \jdeq [ \sigma ]_{\Ty}\;([ \tau ]_{\Ty}\;A) \\
[ \pi_1(\sigma, t) ]_{\Ty}\,A & \jdeq [\sigma]_\Ty\,A \\
[ \pi_1(\sigma; \tau) ]_{\Ty}\,A & \jdeq [\sigma]_\Ty ([\pi\_1\tau]_\Ty \,A) \\
[ \sigma ]_{\Ty}\,U           & \jdeq U
\end{alignat*}

\begin{alignat*}{3}
  \mathsf{idr}    & : {\sigma ; \idS_{\Delta}} && =^{\Sub\,\Gamma\,\Delta} && {\sigma} \\
  \mathsf{idl}    & : {\idS_{\Gamma} ; \sigma} && =^{\Sub\,\Gamma\,\Delta} && {\sigma} \\
  ;\text{-}\mathsf{assoc} & : (\sigma ; \tau) ; \gamma && =^{\Sub\,\Gamma\,\Theta} &&  \sigma ; (\tau ; \gamma) \\
  \mathsf{concat} & : \sigma ; (\tau , t)      && =^{\Sub\,\Gamma\,(\Theta, A)} &&  (\sigma ; \tau) , [ \sigma ] t \\
  \emptyctx\eta   & : \sigma                   && =^{\Sub\,\Gamma\,\emptyctx} & \emptysub \\
  \pi\eta         & : \sigma                   && =^{\Sub\,\Gamma\,(\Delta, A)} &&  (\pi_1 \sigma, \pi_2 \sigma) \\
  \pi_1\beta      & : \pi_1(\sigma , t)        && =^{\Sub\,\Gamma,\Delta} &&  \sigma \\
  \pi_2\beta      & : \pi_2(\sigma , t)        && =^{\Tm\,\Gamma\, A} &&  t \\
  [\idS]t         & : {[\,\idS\,]\,t}          && =^{\Tm\,\Gamma\,A} && t \\
  [;]t            & : {[\,\sigma ; \tau\,]\,t} && =^{\Tm\,\Gamma\,[\sigma ; \tau] A} && {[ \sigma ]\,[ \tau ]\,t} \\
\end{alignat*}

\subsection{... with a Coquand universe} \label{subsec:SC+U}
\cite{Coquand2013}
\begin{alignat*}{3}
[ \idS ]_{\Tm}\,t          & = t \\
[ \sigma ; \tau ]_{\Tm}\,t & = [ \sigma ]_{\Tm}\;[ \tau ]_{\Tm}\;t \\
[ \pi_1(\sigma, t) ]_{\Tm}\,t & \jdeq [\sigma]_\Tm\,t \\
[ \pi_1(\sigma; \tau) ]_{\Tm}\,t & \jdeq [\sigma]_\Tm ([\pi\_1\tau]_\Tm \,t) \\
[ \sigma ]_{\Ty}\,U           & \jdeq U
[ \sigma ]_{\Tm}\,t        & = [ \sigma ]\,t & \text{otherwise}  \\
\end{alignat*}
\subsection{... and \texorpdfstring{$\Pi$}{Π}-types} \label{subsec:SC+U+Pi}

\subsection{... and other type formers} \label{subsec:SC+U+Pi+more}

\section{Eliminators and their examples}
\begin{enumerate}
  \item The elimination rule is somewhat surprising

\end{enumerate}

\section{Equivalence to type theories as quotient inductive-inductive types}
\section{Metatheoretic properties}

@@

\paragraph*{Formalising definitions by overlapping patterns}

\paragraph*{A general schema of quotient inductive-inductive-recursive types}

>>>>

\usepackage{todonotes}
\usepackage{manfnt}
\newcommand{\danger}{\marginpar[\hfill\dbend]{\dbend\hfill}}
\newcommand{\LT}[1]{\todo[inline,author={L-T},caption={}]{#1}}

@@

\begin{abstract}
  Type theory can be defined in a type theory as a quotient inductive-inductive type, but terms are littered with explicit coercions, i.e.\ \transp's, along equality constructors and proofs needs to take account of coercions painstakingly alongside the interesting part.
  This mess has been dubbed the `transport hell' but typically suppressed in presentation for clarity, hiding the gap between the intention and the formalisation of type theory in type theory.
  In this paper, we aim for shortening the gap using quotient inductive-inductive-recursive types and definitions by overlapping patterns altogether, reducing the use of \transp's.
  As a case study, we investigate (parallel) substitution calculus and type theory with Coquand universes and $\Pi$-types as quotient inductive-inductive-recursive types and compare ours with quotient inductive-inductive definitions.
\end{abstract}

\section{Introduction} \label{sec:intro}
\LT{%
Points to elaborate:
\begin{enumerate}
  \item QIITs are littered with transports, causing too much pains while formalising a type theory.
  \item Rewrite rules are popular to make structural rules definitional but requires efforts to justify meta-theoretic properties in general.
  \item Further, the confluence check adopted by \Agda is based on complete development and requires one-step parallel reductions to have a confluent term, making a definition more complicated than necessary.
  \item Moreover, the use of rewrite rules is typically \emph{not} presented and discussed in literature. We are ignoring the elephant in the room.
  \item Instead, definitions by overlapping patterns (in theory) requires strong normalisation and local confluence, making definitions simpler to design.
  \item The rules for type substitutions are \emph{structural} shared with other type theories based on parallel substitution (in particular cwf).

\end{enumerate}
}

\paragraph*{Contributions}
\begin{itemize}
  \item Exploration of the use of quotient inductive-inductive-recursive types and definitions by overlapping patterns.
    In particular, we give in a type theory:
    \begin{itemize}
      \item a definition of parallel substitution calculus;
      \item a definition of type theory with Coquand universes and $\Pi$-types,
    \end{itemize}
   using quotient inductive-inductive-recursive types and definitions by overlapping patterns to define type substitution and term substitution partially.
  \item Comparison with other definitions using quotient inductive-inductive types.
\end{itemize}

@@

\section{Metatheory and formalisation}
\cite{UFP2013}
\LT{
Points to include:
\begin{enumerate}
  \item identity type $x =^{A} y$ for $x, y : A$, dependent identity type, $t =^{P}_{p} u \defeq \transfib{P}{p}{t} =^{P y} u$ for $t : P(x)$ and $u : P(y)$, heterogeneous equality~\cite{McBride1999} in \Agda $x \simeq y$ for $x : A$ and $y : B$
  \item We work with intensional type theory with uniqueness of identity proof and function extensionality (only used for NbE and the standard model for the extensional identity type).
  \item We use Agda to formally implement our definitions with the following options: \texttt{-{}-with-K}, \texttt{-{}-local-confluence-check}, \texttt{-{}-exact-split}, and \texttt{-{}-rewriting} using postulated equations~\cite{Licata2011} to introduce equality constructors for quotient types.
\end{enumerate}}

@@

\LT{Mention that the recursion part of induction-recursion are already not defined by elimination rule.}

\subsection{Definitions by overlapping patterns}
\cite{Cockx2014,Altenkirch2016a}
\subsection{Quotient inductive(-inductive)-recursive types}
\LT{Give an example of QIRT: join list?}
\subsection{Definitions by rewrite rules}
\cite{Cockx2020,Cockx2021}

@@

\section{Type theories as quotient inductive-inductive-recursive types} \label{sec:QIIRTs}
\LT{
\begin{enumerate}
  \item Goal: no transports in the definition.
  \item Motivation: make type substitution definitional so that we do not have to apply transport along structural rules for types explicitly.
  \item Develop the quotient inductive-inductive-recursive definition of type theory step by step (which needs to be locally confluent and terminating).
    \begin{enumerate}
      \item parallel substitution: type substitution
      \item the type of elements of $A$: term substitution needs to be split into two definitional substitution and explicit term substitution to maintain local confluence.
      \item $\Pi$-type: because of the substitution lifting $\sigma^+ \defeq (\pi_1\idS, \pi_2\idS)$ is used we need to introduce a definitional lifting to maintain local confluence.
        We prefer categorical combinator and avoid using $\left< t \right>$ in the definition, which would introduce transports in the definition.
      \item Other type formers can be introduced as usual (as long as we use categorical combinators).
    \end{enumerate}
\end{enumerate}
}
\subsection{Parallel substitution calculus} \label{subsec:SC}
\begin{alignat*}{3}
  \Ctx   & : \Set                   \\
  \Ty    & : \Ctx \to \Set          \\
  \Sub   & : \Ctx \to \Ctx \to \Set \\
  \Tm    & : (\Gamma : \Ctx) \to \Ty\,\Gamma \to \Set
\end{alignat*}
\begin{alignat*}{3}
  \emptyctx & : \Ctx \\
  \blank,\blank & : (\Gamma : \Ctx) \to \Ty\,\Gamma \to \Ctx
\end{alignat*}
\begin{alignat*}{3}
  [\blank]\blank &: \implicit{\Gamma, \Delta} \; \Sub\,\Gamma\,\Delta \to \Ty\,\Delta \to \Ty\,\Gamma \\
\end{alignat*}

\begin{alignat*}{3}
  \UU & : \implicit{\Gamma} & \Ty\, \Gamma
\end{alignat*}
\begin{alignat*}{3}
  \emptysub & : \implicit{\Gamma}\;\Sub\,\Gamma\,\emptyctx \\
  \blank,\blank & : \implicit{\Gamma, \Delta, A}\;(\sigma : \Sub\,\Gamma\,\Delta) \to \Tm\,\Gamma\,([ \sigma ]_{\Ty} A) \to \Sub\,\Gamma,(\Delta, A) \\
  \idS & : \implicit{\Gamma}\;\Sub\,\Gamma\,\Gamma \\
  \blank;\blank & : \implicit{\Gamma, \Delta, \Theta}\;\Sub\,\Gamma\,\Delta \to \Sub\,\Delta\,\Theta \to \Sub\,\Gamma\,\Theta \\
  \pi_1 & : \implicit{\Gamma, \Delta, A}\;\Sub\,\Gamma\,(\Delta, A) \to \Sub\,\Gamma\,\Delta \\
  \pi_2 & : \implicit{\Gamma, \Delta, A}\;(\sigma : \Sub\,\Gamma\,(\Delta, A)) \to \Tm\,\Gamma\,([ \pi_1\,\sigma ]\, A) \\
  [\blank] \blank & : \implicit{\Gamma,\Delta, A}\;(\sigma : \Sub\,\Gamma\,\Delta) \to \Tm\,\Delta\,A \to \Tm\,\Gamma\,([\sigma]\, A)
\end{alignat*}

@@

\begin{alignat*}{5}
  \mathsf{idr}    & : \implicit{\Gamma, \Delta, \sigma} && {\sigma ; \idS_{\Delta}} && =^{\Sub\,\Gamma\,\Delta} && {\sigma} \\
  \mathsf{idl}    & : \implicit{\Gamma, \Delta, \sigma} && {\idS_{\Gamma} ; \sigma} && =^{\Sub\,\Gamma\,\Delta} && {\sigma} \\
  ;\text{-}\mathsf{assoc} & : \implicit{\Gamma, \Delta, \Theta, \Xi, \sigma, \tau, \gamma} && (\sigma ; \tau) ; \gamma && =^{\Sub\,\Gamma\,\Theta} &&  \sigma ; (\tau ; \gamma) \\
  \mathsf{concat} & : \implicit{\Gamma, \Delta, \Theta, \sigma, \tau, A, t} && \sigma ; (\tau , t)      && =^{\Sub\,\Gamma\,(\Theta, A)} &&  (\sigma ; \tau) , \alert{\transfib{\Tm\,\Gamma}{[;]_{\Ty}^{-1}}{\color{black}[ \sigma ] t}} \\
  \emptyctx\eta   & : \implicit{\Gamma, A, \sigma} && \sigma                   && =^{\Sub\,\Gamma\,\emptyctx} & \emptysub \\
  \pi\eta         & : \implicit{\Gamma, \Delta, \sigma} && \sigma                   && =^{\Sub\,\Gamma\,(\Delta, A)} &&  (\pi_1 \sigma, \pi_2 \sigma) \\
  \pi_1\beta      & : \implicit{\Gamma, \Delta, \Theta, \sigma, A, t} && \pi_1(\sigma , t)        && =^{\Sub\,\Gamma,\Delta} &&  \sigma \\
  \pi_2\beta      & : \implicit{\Gamma, \Delta, \Theta, \sigma, A, t} && \pi_2(\sigma , t)        && =^{\Tm\,\Gamma\,([\blank]\,A)}_{\alert{\pi_1\beta}} &&  t \\
  [\idS]_t         & : \implicit{\Gamma, A, t} && {[\,\idS\,]\,t}          && =^{\Tm\,\Gamma}_{\alert{[\idS]_\Ty}}  && t \\
  [;]_t            & : \implicit{\Gamma, \Delta, \Theta, \sigma, \tau, t} && {[\,\sigma ; \tau\,]\,t} && =^{\Tm\,\Gamma}_{\alert{[;]_{\Ty}}}   && {[ \sigma ]\,[ \tau ]\,t} \\
\end{alignat*}

@@

  To better illustrate the annoyance, note that a complete proof of \eqref{eq:pi2-comp-real-proof} requires us to show each of following equations:
  \begin{alignat*}{3}
         & ([\pi_1(\sigma; \tau)]\,A                    &&, \pi_2\,(\sigma ; \tau)) \label{eq:pi2-proof-1} \\
    = {} & ([\pi_1(\sigma; (\pi_1\tau , \pi_2\tau))]\,A &&, \pi_2\,(\sigma; (\pi_1\,\tau, \pi_2\,\tau))) \\
    = {} & ([\pi_1(\sigma;\pi_1\,\tau, t)]\,A &&, \pi_2\,(\sigma;\pi_1\,\tau, t)) \\
    = {} & ([\pi_1(\sigma;\pi_1\,\tau, t)]\,A &&, \pi_2\,(\sigma;\pi_1\,\tau, t)) \nonumber \\
    = {} & (\sub{\sigma;\pi_1\tau}{A} &&, [\sigma] (\pi_2\,\tau)) \nonumber
  \end{alignat*}
  where $p \judgeq [;]_{\Ty}^{-1} : \sub{\sigma}{\sub{\pi_1\,\tau}{A}} \to \sub{\sigma; \pi_1 \tau}{A}$ and $t \judgeq\transfib{\Tm\,\Gamma}{p}{[\sigma]\,(\pi_2\,\tau)}$,
\end{example}
explicitly.
The first three equations and the last equation follow from Lemma~2.3.4 and 2.3.2 in \cite{UFP2013} respectively, which cannot be proved naively as we did in \eqref{eq:pi2-comp-proof}.

\begin{alignat*}{3}
  [\blank]_{\Ty} \blank &: \implicit{\Gamma, \Delta} \; \Sub\,\Gamma\,\Delta \to \Ty\,\Delta \to \Ty\,\Gamma \\
[ \idS ]_{\Ty}\,A             & \jdeq A \\
[ \sigma ; \tau ]_{\Ty}\,A    & \jdeq [ \sigma ]_{\Ty}\;([ \tau ]_{\Ty}\;A) \\
[ \pi_1(\sigma, t) ]_{\Ty}\,A & \jdeq [\sigma]_\Ty\,A \\
[ \pi_1(\sigma; \tau) ]_{\Ty}\,A & \jdeq [\sigma]_\Ty ([\pi_1\tau]_\Ty \,A) \\
[ \sigma ]_{\Ty}\,\UU           & \jdeq \UU
\end{alignat*}

\begin{alignat*}{5}
  \mathsf{concat} & : \implicit{\Gamma, \Delta, \Theta, \sigma, \tau, A, t} &&\sigma ; (\tau , t)      && =^{\Sub\,\Gamma\,(\Theta, A)} &&  (\sigma ; \tau) , [ \sigma ] t \\
  \pi_2\beta      & : \implicit{\Gamma, \Delta, \Theta, \sigma, A, t} && \pi_2(\sigma , t)        && =^{\Tm\,\Gamma\, A} &&  t \\
  [\idS]t         & : \implicit{\Gamma, A, t} && {[\,\idS\,]\,t}          && =^{\Tm\,\Gamma\,A} && t \\
  [;]t            & :\implicit{\Gamma, \Delta, \Theta, \sigma, \tau, t} && {[\,\sigma ; \tau\,]\,t} && =^{\Tm\,\Gamma\,[\sigma ; \tau] A} && {[ \sigma ]\,[ \tau ]\,t} \\
\end{alignat*}

\begin{proposition}[Local confluence]
  Type substitution $[\blank]_{\Ty}\blank$ is locally confluent.
  \danger
\end{proposition}
\begin{proposition}[Coherence]
  $[\sigma]_{\Ty}\,A$ is (propositionally) equal to $[\tau]_{\Ty}\,A$ whenever $\sigma = \tau$.
  \danger
\end{proposition}

\subsection{... with a Coquand universe} \label{subsec:SC+U}
\cite{Coquand2013}
\begin{alignat*}{3}
  \UU  & : \implicit{\Gamma}\;\Ty\, \Gamma \\
  \El & : \implicit{\Gamma}\;\Tm\,\Gamma\;\UU \to \Ty\,\Gamma \\
  [ \sigma ]_{\Ty}\,(\El\, u) & \jdeq \El\,([\sigma]_{\Tm}{u}) \\
\end{alignat*}
\begin{alignat*}{3}
  [\blank]_{\Tm}\blank & : (\sigma : \Sub\,\Gamma\,\Delta) \to \Tm\,\Delta\,A \to \Tm\,\Gamma\,([\sigma]_{\Ty}\,A) \\
[ \idS ]_{\Tm}\,t          & \jdeq t \\
[ \sigma ; \tau ]_{\Tm}\,t & \jdeq [ \sigma ]_{\Tm}\;([ \tau ]_{\Tm}\;t) \\
[ \pi_1(\sigma, t) ]_{\Tm}\,t & \jdeq [\sigma]_\Tm\,t \\
[ \pi_1(\sigma; \tau) ]_{\Tm}\,t & \jdeq [\sigma]_\Tm\; ([\pi_1\tau]_\Tm \,t) \\
[ \sigma ]_{\Tm}\,t        & \jdeq [ \sigma ]\,t, \quad \text{otherwise}  \\
\end{alignat*}

\begin{proposition}[Local confluence]
  Type and term substitutions $[\blank]_{\Ty}\blank$ and $[\blank]_{\Tm}\blank$ are locally confluent.
  \danger
\end{proposition}

\begin{proposition}[Coherence]
  The following statements hold:
  \danger
  \begin{enumerate}
    \item $[\sigma]_{\Ty}\,A = [\tau]_{\Ty}\,A$ whenever $\sigma = \tau$.
    \item $[\sigma]_{\Tm}\,A = [\tau]_{\Tm}\,A$ whenever $\sigma = \tau$.
  \end{enumerate}
\end{proposition}
\subsection{... and \texorpdfstring{$\Pi$}{Π}-types} \label{subsec:SC+U+Pi}
\begin{alignat*}{3}
  \Pi     &: \implicit{\Gamma}            && (A : \Ty\,\Gamma) \to \Ty\,(\Gamma, A) \to \Ty\,\Gamma \\
  \lambda &: \implicit{\Gamma, A, B}      && \Tm\,(\Gamma, A)\,B \to \Tm\,\Gamma\,(\Pi\,A\,B) \\
  \mathsf{app} &: \implicit{\Gamma, A, B} && \Tm\,\Gamma\,(\Pi\,A\,B) \to \Tm\,(\Gamma, A)\,B \\
\end{alignat*}
\begin{alignat*}{3}
  \blank^+ & : \implicit{\Gamma, \Delta, A}\; (\sigma : \Sub\,\Gamma,\Delta) \to (A : \Ty\,\Delta) \to \Sub\,(\Gamma, [\sigma]_{\Ty}\,A)\, (\Delta, A) \\
  \sigma^+ & \defeq (\pi_1\idS ; \sigma, \pi_2\idS) \\
\end{alignat*}
\begin{alignat*}{3}
  \blank\uparrow \blank & : (\sigma : \Sub\,\Gamma,\Delta) \to (A : \Ty\,\Delta) \to \Sub\,(\Gamma, [\sigma]_{\Ty}\,A)\, (\Delta, A) \\
\idS                \uparrow A & \jdeq \idS \\
\sigma ; \tau       \uparrow A & \jdeq (\sigma \uparrow \sub{\tau}{A} ) ; (\tau \uparrow A) \\
\pi_1(\sigma, t)    \uparrow A & \jdeq \sigma \uparrow A \\
\pi_1(\sigma; \tau) \uparrow A & \jdeq \sigma \uparrow (\sub{\pi_1 \tau}{A}) ; (\pi_1 \tau \uparrow A) \\
\sigma              \uparrow A & \jdeq \sigma^+ , \quad \text{otherwiese} \\
  [ \sigma ]_{\Ty}\,(\Pi\,A\,B) & \jdeq \Pi\,(\sub{\sigma}{A})\,(\sub{\sigma\uparrow A}{B})
\end{alignat*}

\begin{alignat*}{3}
  \mathsf{[]\lambda} & : [\sigma]_{\Tm} (\lambda\,t) && =^{\Tm\,\Gamma\,([\sigma]_{\Ty}(\Pi\,A\,B))} && \lambda\,([\sigma \uparrow A]_{\Tm}\,t) \\
  \Pi\beta & : \mathsf{app}(\lambda\,t) && =^{\Tm\,\Gamma,A} {} && t \\
  \Pi\eta & :  t && =^{\Tm\,\Gamma,A} {} && \lambda(\mathsf{app}\,t)
\end{alignat*}

%% ⟨_⟩ : Tm Γ A → Sub Γ (Γ , A)
%% ⟨ t ⟩ = idS , t
%% 
%% _$$_
%%   : (t : Tm Γ (Π A B)) (u : Tm Γ A)
%%   → Tm Γ ([ ⟨ u ⟩ ] B)
%% t $$ u = [ ⟨ u ⟩ ]t app t
\begin{alignat*}{3}
  \left<\blank\right> & : \Tm\,\Gamma\,A \to \Sub\,\Gamma\,(\Gamma, A) \\
  \left< t \right> & \defeq (\idS , t) \\
  \blank \$\blank & : \Tm\,\Gamma\,(\Pi\,A\,B) \to (u : \Tm\,\Gamma\,A) \to \Tm\,\Gamma\,([ \left< u \right> ]\,B) \\
  t \mathop{\$} u & \defeq [ \left< u \right> ]_{\Tm} (\mathsf{app}\,t)
\end{alignat*}

\begin{proposition}
  For every substitution $\sigma : \Sub\,\Gamma\,\Delta$, type $A : \Ty\,\Delta$, and term $t : \Tm\,\Delta\,A$, the following statements hold:
  \danger
  \[
    \sigma \uparrow A = (\pi_1 \idS ; \sigma , \pi_2 \idS)
  \]
  and
  \[
    [ \sigma ]_{\Tm}\,t = [ \sigma ]\,t
  \]
\end{proposition}

\begin{proposition}
  The following statements hold:
  \danger
  \begin{enumerate}
    \item $\sigma \uparrow A = \tau \uparrow A$ whenever $\sigma = \tau$.
    \item $[\sigma]_{\Ty}\,A = [\tau]_{\Ty}\,A$ whenever $\sigma = \tau$.
    \item $[\sigma]_{\Tm}\,t = [\tau]_{\Tm}\,t$ whenever $\sigma = \tau$.
  \end{enumerate}
\end{proposition}
\subsection{... and other type formers} \label{subsec:SC+U+Pi+more}
\LT{%
\begin{enumerate}
  \item Discuss the unit type.
  \item Discuss the Boolean type.
  \item Discuss the type of natural numbers.
  \item Discuss the extensional identity type.
\end{enumerate}
}

\section{The eliminator and the standard model}
\LT{%
\begin{enumerate}
  \item Introduce motives and methods for the definition in \Cref{subsec:SC+U+Pi}.
  \item The rule $\mathsf{Elim}_{\Ty}\,([ \sigma ]_{\Ty}\,A) = [ \mathsf{Elim}_{\Sub}\,\sigma ]^{A}_{\Ty}(\mathsf{Elim}_{\Ty}\,A)$ is not defined but provable. 
  \item $\mathsf{Elim}_{\Tm}\,([ \sigma ]_{\Tm}\,t) = [ \mathsf{Elim}_{\Sub}\,\sigma ]^{A}_{\Tm}(\mathsf{Elim}_{\Tm}\,t)$ is proved.
\end{enumerate}
}

\section{Equivalence to the quotient inductive-inductive definition of type theory}
\LT{%
\begin{enumerate}
  \item Show that two definitions are equivalent (ongoing).
  \item Show that methods are equivalent (ongoing).
\end{enumerate}
}

@@

\paragraph*{A general schema of quotient inductive-inductive-recursive types}
\subparagraph*{Formalising definitions by overlapping patterns}
